\section{Insights, Discussion and Future Prospects}
\label{sec:discussion}
In practice, the three pillars operate together, as the coding-agent example in \S\ref{sec:taxonomy} illustrates. \removed{Four}\added{Five} cross-cutting \removed{challenges}\added{themes} span them: \added{algorithmic interactions and practitioner guidance (\S\ref{sec:algo_interactions}),} feedback-model quality (\S\ref{sec:Feedback_models}), tool-interface design (\S\ref{sec:designing_tools}), adversarial robustness (\S\ref{sec:adversarial_verbal_feedback}), and the absence of formal guarantees (\S\ref{sec:theoritical_foundation}).

\subsection{\texorpdfstring{\added{Algorithmic Interactions and Practitioner Guidance}}{Algorithmic Interactions and Practitioner Guidance}}
\label{sec:algo_interactions}
\added{Different feedback forms pair with different optimization families, and the pairing carries assumptions a practitioner must check. Executable reward functions generated from language pair with online policy-gradient training; pairwise judgments pair with offline preference objectives such as DPO; filtered or feedback-conditioned text pairs with supervised fine-tuning; and step-level verbal judgments pair with process reward models (PRMs), which enable finer-grained credit assignment \citep{lightman2024let}. Each pairing has consequences inside the decision problem: rewards specified through language inherit objective mis-specification and reward-hacking risks; states represented through language raise information-loss and aliasing questions; and actions generated under LLM priors change exploration. Appendix~\ref{app:guide} distills these considerations into a decision table keyed to properties a practitioner can assess before committing to a mechanism: verifiability of task outcomes, feedback latency and cost, and risk exposure such as reward hacking, sycophancy, and feedback poisoning.}


% \das{overall the main shortcoming of this section is the extremely sparse number of papers considering this is the insights section, and even the ones that are there are oldish }

% \das{we also claim earlier is that our contribution is to provide insights and design recommendations, but we are ONLY discussing insights, the section proposes NO recommendations or solutions or hypothesis or empirical recommendations}

% \das{every of sub sections reads like the following: here is an important problem, here is a paper or couple papers, we agree solving the problem is a challenge.}

% \das{do we want to say something like in production all three pillars are increasingly becoming useful together?}




\subsection{Dedicated Feedback Models}
\label{sec:Feedback_models}
Verbal feedback quality remains the key performance factor, but current practice typically reuses the same LLM for both generation and critique. Dedicated critics consistently outperform this approach: Shepherd \citep{wang2023shepherd} provides more actionable feedback than untuned baselines, CriticGPT \citep{mcaleese2024criticgpt} detects code bugs more reliably, and CritiqueLLM \citep{ke2024critiquellm} scales critique data via multi-path prompting. Co-training critics with reasoners yields further improvements \citep{liu2026generativeadversarialreasonerenhancing}, while generative process reward models that reason before scoring preserve more verbal signal than scalar PRMs \citep{khalifa2025processrewardmodelsthink}. CriticBench \citep{lin2024criticbench} offers the first systematic benchmark for critique quality. \added{A broader evaluation practice is emerging around reward-model benchmarks \citep{lambert2024rewardbench} and step-level process benchmarks \citep{zheng2024processbench}; we recommend reporting feedback \emph{quality} and feedback \emph{utilization} separately, since models often fail to incorporate even accurate feedback \citep{jiang2025feedback}.} These results indicate a natural next step: just as the field invested in instruction-tuned models \citep{ouyang2022training}, it should now develop feedback-tuned models whose objectives address error localization, actionability, and calibration. Recent advances in span-level feedback gradients \citep{wang2026text2gradreinforcementlearningnatural} and multi-turn didactic interactions \citep{klissarov2026improving} suggest such models can serve as differentiable training signals across each pillars.




% \subsection{Verbal Feedback Structure}
% The structure and specificity of verbal feedback matter more than its length. \citet{madaan2023selfrefine} found that effective refinement requires critique identifying specific failure points; \citet{thakkar2025can} confirmed this at scale in peer review. Span-level feedback alignment \citep{wang2025text2grad} achieves finer-grained updates than scalar RL, and free-form feedback can serve as a first-class conditioning signal without compression \citep{luo2025verbal}. Early work showed that retaining versus discarding critique text produces measurably different outcomes \citep{scheurer2023training}. Benchmarks such as FeedbackEval \citep{feedbackeval2025} and CriticBench \citep{lin2024criticbench} begin to standardize evaluation.
 
% \textbf{Recommendation.} The field should develop a \emph{feedback specification schema}---structured fields such as \texttt{\{error\_type, location, severity, fix, confidence\}}---enabling systematic comparison of feedback quality independent of downstream task performance.
 


\subsection{Designing Tools for Verbal Feedback}
\label{sec:designing_tools}
Developer-oriented tools often yield outputs that are terse or conflate error types. \citet{yang2024swe} demonstrated that LLM-optimized interfaces enhance agent performance, while \citet{bandlamudi2025framework} showed ambiguous outputs cause many agent failures across 750 API calls. Tool-interactive critiquing requires interpretable output \citep{gou2024critic}, and executable code actions benefit agents by providing structured feedback \citep{wang2024executable}. RL on execution feedback \citep{gehring2024rlef} presumes adequate tool-output signal, while agent harness frameworks \citep{ning2026code} set tool quality as the limit of iterative refinement. However, a metric to measure this limit is lacking: tool-output compliance should quantify agent consumability. Tool APIs must expose structured metadata distinguishing task-level errors from infrastructure failures to enable agents to apply suitable recovery strategies.





\subsection{Adversarial Verbal Feedback}
\label{sec:adversarial_verbal_feedback}
Because VRL agents inherently act on feedback, adversarial feedback constitutes policy manipulation, introducing vulnerabilities beyond standard prompt injection. \citet{zhan2025adaptive} demonstrated that adaptive attacks circumvent all eight tested defenses, while tool-level injection attains a 96.7\% success rate on GPT-4o \citep{shi2025prompt}. Hidden instructions in API responses covertly manipulate agents \citep{zhan2024injecagent}, adversarial memory injection induces cross-session drift \citep{dong2026memory}, and even benign incorrect feedback leads to sycophancy \citep{sharma2024towards}. Defenses are developing: instruction hierarchies \citep{wallace2024instruction}, adversarial fine-tuning \citep{chen2025secaligndefendingpromptinjection}, and system-level design patterns \citep{beurer2025design} each mitigate narrow attack surfaces. Comprehensive solutions will require feedback-provenance mechanisms to verify source integrity and assign trust, as well as adversarial feedback benchmarks to evaluate robustness as the primary metric.
 
% \das{we have to say how this is relevant, is it because something like verbal RL agents act on feedback by design, they're structurally more vulnerable than models that merely generate text becuas of things like prompt injection? }


\subsection{Theoretical Foundations}
\label{sec:theoritical_foundation}
Verbal RL lacks formal guarantees; it remains unclear when verbal feedback enhances sample efficiency or policy quality. Three promising directions emerge. First, if verbal critics serve as noisy oracles with bounded error, PAC-learning frameworks \citep{strehl2009reinforcement} can bound VRL sample complexity, and critic benchmarks \citep{lin2024criticbench} allow empirical estimation of error rates. Second, modeling feedback as partial observation links VRL to POMDP instruction following \citep{luketina2019survey}, where belief-state planning yields convergence guarantees; \citet{mccallum2023feedback} implement this with feedback-conditioned decision transformers. Third, rate-distortion theory may quantify signal retention from critique to scalar, clarifying when preference shaping aligns with feedback-conditioned methods. Empirical studies support these directions: language models learn from verbal feedback without scalar rewards \citep{luo2025fcp}, text feedback broadens RL capabilities \citep{song2026expanding}, and recent surveys link RLHF to language-conditioned RL theory \citep{kaufmann2023survey}. Formalizing results in these areas would clarify when the cost of rich verbal feedback is warranted over scalar alternatives. \added{We state these as open problems rather than established results. Concretely, the field's research agenda includes: the sample-efficiency conditions under which verbal feedback beats scalar reward; optimality and convergence when rewards are specified or shaped through language; temporal credit assignment from free-form critiques; provenance and robustness of feedback channels; and standardized, separate evaluation of feedback quality and feedback utilization. Each item is currently blocked by the absence of a formal account of what a verbal signal contributes beyond its scalarization.}

% \das{this is handwavy for me, why is and how is PAC learning going to help? how is POMDP going to help, we just hypothesized that they will help, but what in the technique there is that will help ? should we talk about convergence in addition to learning dynamics and sample efficiency?}