\section{Introduction}
\label{sec:intro}

\todonote{FIGURE: introduction overview showing the three signal roles (grounding signal, deliberative feedback, learning signal), each annotated with where the signal acts and what it modifies at the moment it takes effect; adapt the ACL main figure but remove the lifecycle-loop arrows (Q27s-3).}

Classical reinforcement learning couples an agent with a well-specified environment, an action space, and a task-specific reward function, and this coupling has produced landmark systems: Atari play from raw pixels \citep{mnih2015human}, championship Go \citep{silver2016mastering}, chip placement \citep{mirhoseini2021graph}, and tokamak plasma control \citep{degrave2022magnetic}. Each success, however, rests on heavy domain-specific engineering: the environment must be formalized, the interaction space fixed, and the reward function must capture the intended behavior closely enough that optimizing it yields what the designer meant \citep{vamplew2021scalarrewardenoughresponse}. Writing such specifications is a long-standing bottleneck \citep{amodei2016concrete}, and it binds most tightly exactly where modern agents are deployed: tasks whose goals are ambiguous, context-dependent, and difficult to encode as a scalar objective.

Large language models \citep{brown2020language, achiam2023gpt} have opened an alternative supervisory channel: natural language itself. Instead of fixing all task intent in advance, verbal feedback communicates corrections, preferences, and constraints on the fly, in a form that agents can increasingly interpret and act on. Frontier models can describe environments \citep{wang2023voyager}, identify bugs in code \citep{mcaleese2024criticgpt}, critique generated outputs \citep{madaan2023selfrefine}, and translate stated preferences into training signals \citep{hong2024orpo, meng2024simpo}. Much as self-supervised learning \citep{devlin2019bertpretrainingdeepbidirectional, chen2020simple} widened the range of data that could supervise model development, verbal feedback is widening the range of signals that can supervise agent improvement.

We call the resulting family of methods \textbf{verbal reinforcement learning (VRL)} and define membership by two criteria. First, the feedback is expressed in natural language, whether self-generated, supplied by a human, or produced by an external tool or model, and its linguistic structure is functionally consumed by the improving agent rather than collapsed to a scalar or a preference bit before use. Second, the feedback participates in an improvement loop: it effects an output revision, a memory write, or a parameter update. Both criteria are required, and together they fence VRL off from adjacent paradigms. Plain instruction following fails the second criterion, since no improvement loop exists. Vanilla reinforcement learning from human feedback (RLHF) \citep{christiano2017deep, ouyang2022training} and direct preference optimization (DPO) \citep{rafailov2023direct} satisfy only the second, because the verbal judgment is reduced to a preference bit before learning; we treat them as the scalar endpoint of a compression spectrum rather than as VRL proper (\S\ref{sec:learning}). Language-conditioned reinforcement learning treats language as an input to the policy rather than as feedback on its behavior. The same criteria settle boundary cases (reward-code generation, in which language builds a reward function that later emits scalars, still qualifies) and scope the definition by modality: only the feedback need be textual, so embodied agents with visual observations remain in scope when the corrections they receive are verbal (\S\ref{sec:embodied}). Section~\ref{sec:framework} develops the canonical, formal version of this fence within a language-augmented sequential decision framework.

The term itself has a short but crowded genealogy. \citet{shinn2023reflexion} coined it to describe self-reflective agents that improve by storing and rereading their own verbal self-critiques, a specific instance of what we classify as deliberative feedback. Our usage is deliberately broader, covering any method that meets the two criteria above rather than only the reflective mechanism for which the term was coined; \S\ref{sec:framework} states the generalization precisely. The nearest adjacent coinage, Natural Language Reinforcement Learning \citep{feng2024natural}, recasts individual reinforcement-learning components such as value functions in language space; it is a framework proposal rather than a survey, and in our terms one formal instantiation of the training-stage cell. Farther back, \citet{luketina2019survey} surveyed reinforcement learning informed by language at a time when language could only inform a scalar-reward learner; in the systems reviewed here that relation is inverted, with language serving simultaneously as the policy medium, the feedback, and increasingly the optimizer trace. The term also has an older life in behaviorist psychology, which \S\ref{sec:framework} takes up alongside the formal genealogy.

Early results indicate that the verbal channel is more than a convenience. \citet{ahn2022can} used language to ground high-level instructions in physical affordances, enabling real-world robotic planning. \citet{shinn2023reflexion} reached 91\% pass@1 on HumanEval through verbal self-reflection alone, and \citet{madaan2023selfrefine} showed that iterative verbal critique yields gains across diverse generation tasks. Even in fully scalarized form, human judgments proved potent: \citet{ouyang2022training} trained a 1.3B-parameter model on preference judgments that outperformed the 175B GPT-3 baseline, and a recurring question in this survey is how much additional value survives when the linguistic structure of feedback is retained rather than compressed away. Nor are these isolated demonstrations: arXiv publications referencing ``self-refine,'' ``self-correction,'' ``verbal feedback,'' and ``language-feedback alignment'' have grown from a handful in early 2020 to several hundred today, spanning coding assistants \citep{yang2024swe, jimenez2024swebenchlanguagemodelsresolve}, scientific discovery \citep{m2024augmenting, lu2024ai}, robotics \citep{wang2023voyager}, mathematical reasoning \citep{hatamizadeh2026igrpo, liu2026dual, shi2025ssr}, clinical decision support \citep{huang2025medreflect, zhou2025enhancing}, and education \citep{qian2025dean, ahtisham2026ai}.

These methods differ in mechanism: some operate during inference, others modify training data, and some redefine the task itself. What they share is the use of natural language to guide, refine, and ground agent behavior. We therefore organize the field along a single axis, when a verbal signal takes effect and what it modifies, yielding three pillars: language as \emph{grounding signal}, which defines the task by specifying goals, states, actions, and reward structure; language as \emph{deliberative feedback}, which refines behavior at inference time without parameter updates; and language as \emph{learning signal}, which shapes parameters at training time. The pillars classify signals, not systems: a single agent may consume signals in several regimes at once, and we make no claim that methods traverse the pillars as a lifecycle loop.

This growth has been accompanied by a crowded shelf of surveys, yet none takes the feedback signal itself as its object of study. Prior overviews organize by mechanism, such as correction timing \citep{pan2023automatically, kamoi2024can}, test-time scaling dimensions \citep{zhang2025testtime}, or rollout stage \citep{surana2026generate}; by agent component, such as memory \citep{zhang2025survey, zheng2025lifelong}, cognitive architecture \citep{sumers2023cognitive}, or the component under self-evolution \citep{tao2024survey, gao2025survey, fang2025comprehensive}; by algorithm family, such as DPO \citep{xiao2024comprehensive}, RLHF \citep{kaufmann2023survey, liu2026reinforcement}, or reinforcement-learning technique \citep{wang2024reinforcement, srivastava2025technical, zhang2025landscape}; or by the producer of judgments \citep{gu2024survey, li2024generation, li2024judges, you2026agent}. Reward-centric treatments come closest to our concern \citep{ji2025survey, wu2025sailing, liu2025enhancing, wei2025unifying}, but they presuppose that feedback must become a reward, which is precisely the design decision a signal-centric account interrogates. The nearest precedent in spirit, a formalization of human feedback for natural language generation by format, objective, and use \citep{fernandes2023bridging}, predates the agentic turn, so reward-code grounding, rubric rewards, and experiential memory fall outside its frame. To our knowledge, no existing survey combines a signal-centric organizing axis with a formal treatment of language as signal, coverage that runs from problem definition through inference-time deliberation to training, and a first-class treatment of the safety of the feedback channel. That combination is the gap this article fills.

This article makes four contributions.
\begin{enumerate}
    \item \textbf{A signal-centric formalism.} We define VRL by the two-criterion fence above and develop a language-augmented sequential decision framework that specifies where a linguistic artifact can enter the decision problem (goal, observation, action prior, reward, transition prior, or improvement operator), separating VRL from RLHF, instruction following, and language-conditioned reinforcement learning (\S\ref{sec:framework}). The taxonomy built on this framework assigns each signal by what it modifies at the moment it takes effect (\S\ref{sec:taxonomy}).
    \item \textbf{Decision rules and practitioner guidance.} We give an explicit assignment rule for placing methods, decision rules for hybrid cases such as reward-code generation and experiential memory, and a prescriptive guide that keys the choice among verbal-feedback mechanisms to task verifiability, feedback latency and cost, and risk (\S\ref{sec:taxonomy}, \S\ref{sec:practitioner}).
    \item \textbf{A credit-assignment and optimization synthesis.} We follow verbal signals into the machinery that consumes them: the verbal reward stack that runs from binary verifiers through rubrics and generative reward models to process reward models (PRMs) (\S\ref{sec:rewardstack}); outcome-level, process-level, and token- or span-level credit assignment, together with the pairing of feedback forms and algorithm families (\S\ref{sec:credit}); and optimizers that update prompts, memory, and context in language space rather than weights (\S\ref{sec:optimizers}). Throughout, we state plainly which formal guarantees exist and which do not.
    \item \textbf{Positioning and agenda.} We compare this survey with more than a dozen prior surveys and overview papers along a fixed set of dimensions (organizing axis, feedback types, stages covered, formalism, practitioner guidance, and feedback safety), and we distill a research agenda from the gaps the comparison exposes (\S\ref{sec:practitioner}).
\end{enumerate}

The remainder of the article proceeds as follows. Section~\ref{sec:framework} presents the formal framework and the full definitional fence, and Section~\ref{sec:taxonomy} introduces the taxonomy, its assignment rule, and a worked coding-agent example. Three sections then develop the pillars: Section~\ref{sec:grounding} covers language as grounding signal, including goal, state, and action grounding and reward-code generation; Section~\ref{sec:deliberative} covers language as deliberative feedback, including self-critique, externally grounded critique, multi-agent debate, experiential memory, and evaluator feedback within search; and Section~\ref{sec:learning} covers language as learning signal along a compression spectrum whose scalar endpoint is vanilla RLHF and DPO. Section~\ref{sec:rewardstack} surveys the verbal reward stack, Section~\ref{sec:credit} treats credit assignment and optimization with verbal signals, and Section~\ref{sec:optimizers} covers language-space optimizers. Section~\ref{sec:embodied} extends the account to embodied and multimodal settings, Section~\ref{sec:evaluation} synthesizes benchmarks for feedback quality and utilization, and Section~\ref{sec:safety} examines attacks on the feedback channel and defenses against them. Section~\ref{sec:practitioner} collects the practitioner guide, the comparison with prior surveys, and the research agenda; Section~\ref{sec:conclusion} concludes. Practitioners deciding which mechanism to adopt may read Sections~\ref{sec:taxonomy} and \ref{sec:practitioner} first; readers interested in theoretical questions may proceed from Section~\ref{sec:framework} directly to Section~\ref{sec:credit}.
