\section{Pillar 1: Language as Grounding Signal}
\label{sec:grounding}

The first pillar contains signals that define the decision problem before behavior is evaluated. Language can specify the goal, abstract the state, constrain the action interface, or construct the reward. A method belongs here when the linguistic artifact modifies one of those task components and the resulting component drives an improvement loop. This placement distinguishes grounding from one-shot instruction following and from feedback that revises an existing trajectory. Across all four mechanisms, the binding constraint is not descriptive detail but whether language maps precisely enough to an executable component of the decision process \citep{sutton2018reinforcement,macglashan2017interactive}.

\subsection{Goal Grounding}
\label{subsec:goal-grounding}

Goal grounding turns a verbal specification into target conditions that later optimization or revision can pursue. It must resolve objects and relations, filter plans by physical feasibility, and decompose an open-ended request into checkable subgoals. Systems such as SayCan combine linguistic relevance with affordance estimates, while language-specification methods train policies without a manually written reward \citep{ahn2022can,li2026ground}. The central failure mode is compositional generalization: familiar primitives remain difficult when combined into a novel goal.

\subsection{State Grounding}
\label{subsec:state-grounding}

State grounding uses language as an observation abstraction. Text environments make this role explicit, while embodied agents obtain it by translating visual or physical state into descriptions. Transfer from verbal environments to embodied settings suggests that language can provide a modality-invariant representation, and closed-loop descriptions can support replanning \citep{shridhar2020alfworld,bhat2024grounding}. The relevant test is sufficiency, not fluency: a description fails when it omits history or state variables needed to distinguish actions with different values.

\subsection{Action Grounding}
\label{subsec:action-grounding}

Action grounding maps instructions to skills, tool calls, programs, or motor commands and returns execution outcomes to the planner. Inner Monologue illustrates the closed loop, using verbalized success and scene feedback to revise plans, while ProgPrompt constrains generation with program-like descriptions of available objects and actions \citep{huang2022inner,singh2022progprompt}. The principal design choice is granularity. A phrase may denote a long skill, a discrete subgoal, or a motor adjustment, and that choice fixes both the interface and the horizon over which credit must be assigned.

\subsection{Reward Grounding: From Language to Reward Functions}
\label{subsec:reward-grounding}

Reward grounding compiles a verbal description of success into an executable objective. It is the key boundary case in the definitional fence: linguistic structure is consumed when the reward is constructed, even though the resulting function later emits scalars. Eureka searches over generated reward programs using textual reflection, and Text2Reward produces dense reward code that people can refine \citep{ma2024eureka,xie2024text2reward}. When state is not directly available, learned vision-language judges can instead ground a text goal in observations \citep{wang2024rlvlmf}. These mechanisms fail differently from ordinary reward estimators. A miscompiled reward is a specification error that can be optimized confidently, so fidelity and hacking resistance must be tested before policy performance is trusted.

\subsection{Synthesis: Shared Failure Modes and When to Ground}
\label{subsec:grounding-synthesis}

Three risks recur. The grounding gap separates a fluent description from an executable mapping. Compositionality failures appear when individually understood pieces interact. Fidelity failures arise when a valid task component does not mean what its author intended. Grounding is therefore the right first move when the task is not yet well posed: no programmatic objective exists, observations lack a useful encoding, or the action interface is unsettled. Once language has produced a testable problem, deliberative feedback can revise episodes and learning signals can reshape the policy. Detailed systems, quantitative comparisons, and secondary grounding variants are collected in the electronic supplement.
The four roles also fail on different timescales. Goal and reward errors can contaminate an entire training run, state omissions recur at every observation, and action mismatches surface during execution. Validation should therefore occur at the interface each role constructs, before downstream optimization obscures the original grounding error.
