\section{Pillar 2: Language as Deliberative Feedback}
\label{sec:deliberative}

Deliberative feedback refines behavior within an episode without updating model parameters, the regime closest to test-time compute scaling \citep{snell2024scaling}. The four mechanisms differ mainly in where feedback originates: the acting model or an external tool, peer models, stored experience, or an evaluator inside search. A signal belongs here when its text is consumed in the current context; if the same artifact is converted into training data or weights, it becomes a learning signal. The organizing trade-off is among feedback quality, circularity, and inference cost.

\subsection{Self-Critique and Externally Grounded Critique}
\label{subsec:critique}

Self-critique asks a model to diagnose and revise its own output. Self-Refine established the basic loop, but correction is limited when generator and critic share the same blind spots \citep{madaan2023selfrefine,kamoi2024can}. External grounding changes the evidence available to the critic. Unit tests, execution traces, search results, and tool responses can expose errors that introspection cannot, as CRITIC demonstrates across reasoning and tool-use tasks \citep{gou2024critic}. This shift improves reliability but introduces latency, infrastructure cost, and a new interpretation problem: the tool may report the right symptom while the model infers the wrong cause. Grounded critique is therefore strongest where an inexpensive and auditable oracle exists.

\subsection{Multi-Agent Debate}
\label{subsec:debate}

Debate distributes critique across model instances. Its benefit comes from genuine diversity rather than agent count alone: heterogeneous roles or models can reveal distinct errors, while identical debaters often converge to redundant consensus \citep{du2023debate,zhang2025stop}. Debate can also support scalable oversight when a weaker judge evaluates stronger participants, but it multiplies tokens and model calls. It is most defensible for problems that are hard relative to the base model and for which diverse reasoning paths are available; otherwise self-consistency or a single grounded critic is usually cheaper.
Communication topology matters as much as participant count. Sparse exchange, selective retention of disagreements, and explicit confidence can reduce cost or prevent premature consensus. The relevant baseline is therefore not another debate protocol alone, but the strongest single-agent method at the same inference budget.

\subsection{Experiential Memory}
\label{subsec:memory}

Experiential memory stores linguistic feedback for later episodes. The classification follows consumption: a retrieved lesson that conditions the current context remains deliberative even though the store persists. Reflexion anchors the raw-reflection regime, while ExpeL distills portable lessons from contrasts between successful and failed trajectories \citep{shinn2023reflexion,zhao2024expel}. What should persist depends on recurrence. Raw episodes suit closely repeated tasks, abstract principles transfer across related tasks, and executable procedures suit behavior that must be reproduced reliably. Store management becomes part of the mechanism as experience accumulates: selective writing, retrieval, pruning, and evidence-backed admission prevent stale or incorrect lessons from dominating. Memory differs from a language-space optimizer when the agent writes and reads its own store during episodes; an outer loop that edits a prompt or playbook against a task metric instead treats text as a system parameter.
Because retrieval changes the effective context, memory evaluation must test more than recall. A useful store should improve behavior on later tasks, reject irrelevant lessons, preserve raw evidence when summaries are uncertain, and forget entries whose utility has decayed.

\subsection{Evaluator Feedback in Search}
\label{subsec:search}

Search methods branch over candidate trajectories and use linguistic evaluation to choose which nodes to expand or abandon. Tree of Thoughts and Reasoning via Planning are representative designs \citep{yao2023tree,hao2023reasoning}. Their advantage is early rejection of unproductive paths; their cost is repeated evaluation across many states. Sampling or voting without a verbal evaluator is outside the survey's fence.
Search is preferable when partial trajectories can be judged before completion and when that early pruning saves more computation than the evaluator consumes. Otherwise, evaluator noise can discard the correct branch while adding cost.

\paragraph{Synthesis: when this pillar wins:}
Feedback quality is the first decision variable. Self-critique is cheap but circular, debate helps only when participants contribute distinct evidence, tools are reliable only where a verifier exists, and memory can entrench early mistakes. Compute is the second variable: debate and search spend additional calls within an episode, while memory amortizes prior deliberation across episodes. Deliberative feedback is therefore the preferred mechanism when weights are inaccessible, tasks are difficult enough to justify iteration, or recurring tasks let stored lessons pay for themselves. When training access and a stable task distribution are available, the same critiques are often better internalized as learning signals. Extended system catalogs, benchmark results, memory-management variants, and the full mechanism diagram appear in the electronic supplement.
