\section{The Verbal Reward Stack}
\label{sec:rewardstack}

Training-time VRL depends on how an evaluation becomes the reward consumed by an optimizer. Four constructions form a progression: verifiable outcome rewards, rubric rewards, generative reward models, and process reward models (PRMs). The rungs are ordered by two properties: where linguistic structure collapses into a scalar, which we call the \emph{scalarization boundary}, and the granularity at which the resulting signal attaches to a trajectory. Moving upward retains language later and localizes credit more finely, but increases verification cost. Table~\ref{tab:rewardstack} summarizes the stack.

\begin{table}
\caption{The verbal reward stack, ordered by scalarization boundary and signal granularity.}
\label{tab:rewardstack}
\small
\begin{tabular}{@{}p{0.14\textwidth}p{0.26\textwidth}p{0.22\textwidth}p{0.11\textwidth}p{0.15\textwidth}@{}}
\toprule
Rung & Reward computation & Scalarization boundary & Granularity & Representative work \\
\midrule
Verifiable outcome & Programmatic check against ground truth & At the environment; no linguistic structure in the reward & Episode & \citep{lambert2024tulu,guo2025deepseek} \\
Rubric reward & Judge checks instance-specific criteria; item scores pooled & At aggregation of criterion verdicts & Response & \citep{gunjal2025rubrics,viswanathan2025checklists} \\
Generative reward model & Judge writes principles and critique, then a verdict & At verdict emission; critique retained as an audit trail & Response & \citep{zhang2025genrm,liu2025inferencetime} \\
Process reward & Step-level verification, increasingly verbalized & Per step, or per token at the limit & Step or token & \citep{wang2024mathshepherd,khalifa2025processrewardmodelsthink} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Verifiable Outcome Rewards}
\label{subsec:rlvr}

Reinforcement learning with verifiable rewards uses programmatic correctness checks. Tulu 3 established the open post-training recipe, and DeepSeek-R1 showed that outcome-only reinforcement can elicit verification behavior without labeled reasoning traces \citep{lambert2024tulu,guo2025deepseek}. This is the degenerate rung: scalarization occurs in the environment and no language appears in the reward. Its strengths are exactness and low cost; its limits are sparse feedback, narrow domain coverage, and evaluation sensitivity to compute budgets.
Outcome verification is therefore the default only when correctness is both mechanically testable and difficult for the policy to game. Otherwise it supplies a reliable terminal label but little guidance about how a failed trajectory should change.

\subsection{Rubric Rewards}
\label{subsec:rubrics}

Rubrics extend verification to open-ended tasks by decomposing success into instance-specific criteria. Rubrics as Rewards demonstrates that structured criteria can outperform direct holistic judging, while checklist rewards apply the same idea to instruction following \citep{gunjal2025rubrics,viswanathan2025checklists}. Static banks amortize authoring cost, dynamic criteria adapt to policy drift, and selective checking saves judge calls. In every case, scalarization occurs when criterion verdicts are aggregated, so any omitted requirement becomes invisible to the optimizer.
A rubric should therefore be treated as a versioned specification, with coverage audits and updates tied to new policy failures rather than as a fixed scoring prompt.

\subsection{Generative Reward Models}
\label{subsec:genrm}

Generative reward models verbalize the evaluation before emitting a verdict. Generative verifiers write a rationale and benefit from sampling or voting, while DeepSeek-GRM generates principles and critiques before scoring \citep{zhang2025genrm,liu2025inferencetime}. Their text can improve judgment and provide an audit trail, but the optimizer still receives a scalar unless the critique is separately routed into revision or training. Correct verdicts do not guarantee sound explanations, so critique quality must be supervised or evaluated independently.
Their extra inference cost is justified when hard cases require adaptive reasoning or when the retained critique supports auditing, downstream revision, or later credit assignment.

\subsection{Process Reward Models}
\label{subsec:prm}

PRMs move evaluation inside the trajectory. Math-Shepherd automated step labeling, and ThinkPRM generates a verification rationale for each step \citep{wang2024mathshepherd,khalifa2025processrewardmodelsthink}. Fine granularity can localize the first error and improve credit assignment, but noisy step labels and weak transfer beyond curated reasoning tasks remain central limitations. Generative PRMs connect this rung to verbal feedback because their linguistic reasoning participates in computing each local score.
They are most useful when intermediate steps have stable semantics and can be checked independently; otherwise the apparent precision of a per-step score can exceed the reliability of its annotation.

\subsection{The Scalarization Boundary}
\label{subsec:scalarization}

Every rung ultimately sends numbers to the optimizer. Verifiers scalarize at the environment, rubrics after criterion aggregation, generative judges at verdict emission, and PRMs per step. What disappears is the repair information carried by language: which failure occurred, where it occurred, and what should change. Finer granularity narrows that loss but does not eliminate it; language-space optimizers do so by applying the critique as text.

The boundary is also where exploitation concentrates. A policy can target whatever a verifier, rubric, or judge fails to communicate through its scalar output. Outcome checks invite shortcut reasoning, static rubrics leave exploitable omissions, generative judges can pair correct verdicts with unsound critiques, and noisy step labels corrupt PRM training. The practitioner rule is therefore to choose the lowest-cost rung that captures the distinctions the task requires, then test the scalarization boundary for omitted criteria, judge error, and hacking. Extended reward-model variants, quantitative results, and defense evidence appear in the electronic supplement.
