\section{Language Feedback in Embodied and Multimodal Settings}
\label{sec:embodied}

Embodied systems receive images, video, and proprioception and emit motor actions, so language may appear only in the feedback channel. Here \emph{verbal} denotes the structure of the consumed signal, not its carrier: typed text, transcribed speech, or a vision-language model's explanation qualifies when compositional language changes behavior. A score derived from the same observation does not. Language-conditioned control also remains outside scope when the instruction specifies the task but does not evaluate prior behavior.

\subsection{Four Embodied Roles}
\label{subsec:robot-corrections}

\emph{Corrections} map a phrase to a policy change during execution. LILAC grounds real-time utterances in a low-dimensional control space, illustrating the central challenge: the system must identify both what failed and which motor degree of freedom should change \citep{cui2023no}. Corrections can persist as memory or training data, but their taxonomy assignment follows where the language is consumed.

\emph{Language-to-reward} uses linguistic feedback to define or revise an objective before policy optimization. Eureka verbalizes training statistics and mutates reward code, making reward design itself a language-space optimization loop \citep{ma2024eureka}. This family fits simulators that expose state and optimization, but inherits reward misspecification and hacking risk.

\emph{Vision-language rewards and critics} judge behavior when only pixels are available. RL-VLM-F learns rewards from vision-language preference labels over observations \citep{wang2024rlvlmf}; process variants provide denser temporal credit. Their main risk is partial observability: a judge may recognize objects while missing sequence, motion, or causal progress.

\emph{Failure reasoning} produces an explanation that a planner or controller can use for recovery. AHA detects failures in free-form language and improves downstream planning, verification, and reward refinement when inserted into the loop \citep{duan2024aha}. Unlike a scalar reward, the explanation can identify the failed subtask and suggest a repair, but it still requires grounding back into action.

These roles can coexist in one system without becoming interchangeable. A spoken correction first steers the current action, then becomes memory if stored, and finally becomes a learning signal if logged for retraining. Likewise, a vision-language explanation remains deliberative while it guides replanning, but becomes objective-side supervision when converted into reward code. Classification follows the consumed representation and update locus at each use, not the sensor or model that produced it.

\subsection{Selection Rule}
\label{subsec:embodied-lessons}

Choose human corrections when a person can intervene online; generated rewards when a simulator exposes state; vision-language rewards when observation is pixel-only and scalable supervision is required; and failure critics when autonomous recovery matters. These roles enter different parts of the decision problem: corrections and explanations change the policy-side decision, generated rewards define the objective, and learned visual judges supply a training signal. Across all four, feedback phrasing and diversity are design variables rather than presentation details \citep{xi2024teaching}. The modality changes the grounding burden, not the survey's inclusion rule. Extended real-robot, reward-model, and failure-recovery evidence appears in the electronic supplement.
