\section{Safety and Robustness of Feedback Channels}
\label{sec:safety}

Language is both supervision and an attack surface. Threats differ by who corrupts the signal and where it is consumed: an external adversary can manipulate a judge, observation, or training record; the policy can exploit its evaluator; and preference learning can inherit sycophancy without a deliberate attacker. The decomposition identifies which link of the verbal reward stack must be audited and why static capability evaluation is insufficient.

Downstream authority determines severity. A manipulated leaderboard changes a ranking; the same verdict used for learning changes a policy; an injected observation can trigger an external action. Medical, email, coding, and computer-use agents therefore need stricter controls because a linguistic error can become physical, financial, or privacy harm.

\subsection{Robustness of LLM Judges}
\label{subsec:judge-robustness}

Injected phrases can force an LLM judge to prefer attacker-controlled outputs regardless of substance, corrupting evaluation, search, and training rewards. Optimized attacks outperform manual injections and known-answer or perplexity defenses \citep{shi2024optimization}. Comparative scoring and judge committees reduce exposure, but do not remove transferable artifacts or prompt-template sensitivity.

\subsection{Injection into the Observation Loop}
\label{subsec:observation-injection}

Tool outputs, webpages, and retrieved documents can carry indirect instructions that redirect an agent. InjecAgent demonstrates this attack against tool-integrated ReAct agents \citep{zhan2024injecagent}. The defense boundary is therefore not the model prompt alone: untrusted observations require provenance, privilege separation, and constrained tool execution.

\subsection{Poisoning of Training and Optimizer Feedback}
\label{subsec:feedback-poisoning}

Preference pairs and verbal optimizer feedback can be poisoned upstream. DPO backdoors succeed at lower poisoning rates than PPO-based pipelines in controlled comparisons \citep{pathmanathan2024is}, while language-space optimizers are especially exposed because a manipulated critique is their update direction. Data provenance and feedback-source trust must be reported with performance.

\subsection{Policy Exploitation and Sycophancy}
\label{subsec:policy-hacking}
\label{subsec:sycophancy}

Even a clean channel can be exploited. Interactive feedback loops enable in-context reward hacking that static datasets cannot reveal \citep{pan2024feedback}; underspecified rubrics similarly reward partial or superficial satisfaction. Sycophancy is the systemic counterpart: human and learned preference models can favor convincing agreement over correctness, steering optimization away from truthfulness \citep{sharma2024towards}. Both require behavioral stress tests, not only held-out reward accuracy.

\subsection{Defense Synthesis}
\label{subsec:channel-defenses}

Defenses fall into detection, signal reshaping, and system design. Detection alone is brittle because adaptive attacks optimize against the detector. Signal reshaping hardens what the reward credits, as adversarial verifier training does against master-key responses \citep{zhao2025one}. System design limits what corrupted text can cause through instruction hierarchy, provenance marking, least-privilege tools, and independent checks. Three rules survive across studies: comparative judgments are generally safer than absolute scores; stronger or multiple judges reduce but do not eliminate exploitation; and feedback-channel robustness must be measured under interaction, where attacks and reward hacking compound. A minimum safety report should name the trusted source, attack surface, downstream authority, stress tests, and residual failure mode. The electronic supplement catalogs the supporting attacks, quantitative results, and defenses.

Defense should follow the channel from source to effect. At ingestion, authenticate sources and preserve provenance instead of mixing trusted instructions with untrusted observations. At interpretation, compare independent judgments and test decisions under paraphrase and stylistic changes. At execution, gate high-impact tools, restrict permissions, and confirm irreversible actions. At learning time, inspect poisoning sensitivity and train rewards against high-scoring low-quality outputs. Evaluate the complete loop because reasonable filters, judges, and policies can still compose into an exploitable system. Report the adaptive attacker, utility cost, and authority available after compromise.
