\section{Conclusion}
\label{sec:conclusion}

This article has surveyed verbal reinforcement learning through a single organizing question: where a natural-language signal enters the decision problem and what it modifies at the moment it takes effect, yielding the three pillars of language as grounding signal, as deliberative feedback, and as learning signal. The four contributions promised in \S\ref{sec:intro} were delivered as follows. The two-criterion fence and the language-augmented sequential decision framework of \S\ref{sec:framework} give VRL a definition tight enough to exclude plain instruction following, vanilla reinforcement learning from human feedback, and language-conditioned reinforcement learning, while the taxonomy of \S\ref{sec:taxonomy} assigns every signal by its point of effect and resolves hybrid cases, such as reward-code generation and experiential memory, by explicit decision rules rather than by fiat. These rules are turned outward in the prescriptive practitioner guide of \S\ref{sec:practitioner}, which keys the choice among verbal-feedback mechanisms to task verifiability, feedback latency and cost, and the dominant risk each mechanism carries. The synthesis sections then followed verbal signals into the machinery that consumes them: the verbal reward stack (\S\ref{sec:rewardstack}), credit assignment at outcome, process, and token granularity (\S\ref{sec:credit}), and optimizers that act in language space rather than on weights (\S\ref{sec:optimizers}). Finally, the comparison with prior surveys and the research agenda, also in \S\ref{sec:practitioner}, position this account against its neighbors and state what remains to be done.

The payoff of the signal-centric view is consolidation. Questions ordinarily scattered across separate literatures, such as how much of a critique survives compression to a scalar, which feedback forms pair with which algorithm families, and when a verifier should displace a judge, become instances of one question about signal placement and fidelity. Looking ahead, we expect verbal feedback to become a first-class design primitive in agent architectures. Deployed agents will increasingly consume grounding, deliberative, and learning signals within a single trajectory, and the bottleneck will shift from generating feedback to verifying it; the infrastructure that shift demands, from feedback provenance to standardized quality evaluation to machine-readable tool interfaces, is the substance of the research agenda in \S\ref{sec:practitioner}.

\subsection{Limitations}
\label{subsec:limitations}

The survey's own limits should be stated plainly. Within each category we cover representative papers rather than an exhaustive enumeration, and the taxonomy is an organizing lens rather than a strict partition: several methods span roles, since verbal feedback can serve simultaneously as a grounding signal, a deliberative mechanism, and a learning signal, and our decision rules adjudicate placement without dissolving the underlying overlap. Coverage is restricted to methods in which verbal feedback is explicit and central to the agent's process; adjacent work where language plays an auxiliary role is treated only in passing, and the definitional fence fixes the feedback modality to text, so spoken and richer multimodal feedback channels fall outside our scope even where embodied observations do not (\S\ref{sec:embodied}). The article is also a synthesis: it contributes no new benchmark or controlled empirical comparison, and where the literature offers none we have said so rather than inferred one.

Language-based feedback itself carries failure modes that this survey can name but not yet bound: reward-specification failures when language compiles to the wrong objective, unreliable interpretation of feedback by the consuming model, weak grounding between verbal descriptions and environment state, and uncertain transfer of language-guided policies to real-world settings. Societal impact follows the same channels. When feedback is incorrect, biased, adversarial, or poorly grounded, harm falls on the users of high-stakes deployments, among them coding assistants, robotics, clinical decision support, and education, and on the people affected by those systems' outputs, who rarely chose the feedback channel and cannot audit it. Sections~\ref{sec:safety} and~\ref{sec:credit} discuss mitigation directions and the current state of formal guarantees; narrowing the gap between what the definitional fence admits and what theory can certify is, in our judgment, the field's most consequential open problem.
